---
title: "Modeling"
output: html_document
---

## setup

```{r libraries}
library(car)
library(lme4)
library(lmtest)
library(glmnet)
library(ggplot2)
library(emmeans)
library(sandwich)
library(tidyverse)
library(lubridate)
```

```{r read in data}
df <- read.csv("data/modeling_processed.csv") 
```

### data preparation

```{r}
# convert categorical columns to factors
df$works <- as.factor(df$works)
df$engine <- as.factor(df$engine)
df$Year <- as.factor(df$Year)

# standardize performance within each race and flag extreme outliers
df <- df %>%
  group_by(Year, EventName) %>%   # group by race session (each qualifying event)
  mutate(FastestQ1 = min(Q1_time, na.rm = TRUE),              # fastest lap time in that session
         GapToFastest = Q1_time - FastestQ1,                  # absolute time gap to fastest driver
         PerToFastest = GapToFastest / FastestQ1,             # relative gap (percentage difference)
         is_beyond_107 = ifelse(PerToFastest > 0.07, 1, 0)    # flag laps slower than 107% rule
         ) %>% ungroup()  

# split data out by outliers
df_fast <- df %>% filter(is_beyond_107 == 0)         # within 107%      
df_slow <- df %>% filter(is_beyond_107 == 1)         # beyond 107% 

# center throttle variable
df$Throttle_Centered <- df$FullThrottlePct - mean(df$FullThrottlePct)
```

## modeling

### model 1 (baseline): effect of team status, engine, and year on relative performance gap (excluding very slow drivers)

```{r modeling}
# fit multiple regression model
model_v1 <- lm(PerToFastest ~ works + engine + Year, data = df_fast)
# print regression summary
summary(model_v1)
# anova table
car::Anova(model_v1, type=2) 
```

- model explains ~20.5% of the variation in relative performance gap (Adj. RÂ˛ = 0.205) and model is statistically significant (F-test p < 0.001)
- main effects: works teams have a significantly smaller gap (~0.59 percentage points lower, p < 0.001) and gap decreases over time (~0.095 percentage points per year, p < 0.001)
- engine differences: mercedes associated with a smaller gap (p < 0.01), renault engines associated with a larger gap (p < 0.01), honda effect not statistically significant at 5% level
- anova: works, engine, and year all have statistically significant effects

```{r check assumptions}
par(mfrow = c(2,2))
# get plots
plot(model_v1)              
# normality
shapiro.test(residuals(model_v1))     
# equal variance
car::leveneTest(PerToFastest ~ works * engine, data = df_fast)
# independence
lmtest::dwtest(model_v1)
```
- not normal, not constant variance, not independent

```{r}
coeftest(model_v1, vcov = vcovHC(model_v1, type = "HC1"))
```


### model 2: effect of team status, engine, and year on gap, controlling for throttle

```{r}
# fit model (add throttle as a covariate)
model_v2 <- lm(PerToFastest ~ works + engine + Year + FullThrottlePct, data = df_fast)
# regression summary
summary(model_v2)
# ANCOVA table
car::Anova(model_v2, type = 2)
```

- Adj. RÂ˛ = 0.208 (not much change)
- FullThrottlePct is statistically significant (p < 0.001), with higher throttle usage associated with a slightly smaller performance gap

### model 3 (interaction): year Ă— work Ă— engine effects on gap

```{r}
# fit model
model_v3 <- lm(PerToFastest ~ Year * works * engine, data = df_fast)
# model summary
summary(model_v3)
# ANOVA table
car::Anova(model_v3, type = 2)
```

- Adj. RÂ˛ = 0.313 (improvement over previous models) and is statistically significant.
- interactions: significant two-way interactions (YearĂ—works, YearĂ—engine, worksĂ—engine) and a significant three-way interaction (YearĂ—worksĂ—engine), evidence effect of team status and engine on performance gap changes year over year

```{r pairwise contrasts}
# pairwise engine comparisons within each year x work combination
emmeans(model_v3, pairwise ~ engine | Year * works)
# pairwise works comparisons within each year x engine combination
emmeans(model_v3, pairwise ~ works | Year * engine)
```

- estimated marginal means show in 2022, works teams had a lower relative performance gap than customer teams across most engines
- for customer teams, ferrari and mercedes have higher gaps, while renault has the lowest gap
- for works teams, gaps are generally smaller but vary more by engine

### model comparison

```{r}
AIC(model_v1, model_v2, model_v3)
```

- model 2 is slightly better than model 1 but model 3 is better than both
