---
title: "Modeling"
output: html_document
---

## setup

```{r libraries}
library(car)
library(lme4)
library(lmtest)
library(glmnet)
library(ggplot2)
library(emmeans)
library(sandwich)
library(tidyverse)
library(lubridate)
```

```{r read in data}
df <- read.csv("data/modeling_processed.csv") 
```

### data preparation

```{r}
# convert categorical columns to factors
df$works <- as.factor(df$works)
df$engine <- as.factor(df$engine)
df$Year <- as.factor(df$Year)

# standardize performance within each race and flag extreme outliers
df <- df %>%
  group_by(Year, EventName) %>%   # group by race session (each qualifying event)
  mutate(FastestQ1 = min(Q1_time, na.rm = TRUE),              # fastest lap time in that session
         GapToFastest = Q1_time - FastestQ1,                  # absolute time gap to fastest driver
         PerToFastest = GapToFastest / FastestQ1,             # relative gap (percentage difference)
         is_beyond_107 = ifelse(PerToFastest > 0.07, 1, 0)    # flag laps slower than 107% rule
         ) %>% ungroup()  

# split data out by outliers
df_fast <- df %>% filter(is_beyond_107 == 0)         # within 107%      
df_slow <- df %>% filter(is_beyond_107 == 1)         # beyond 107% 

# center throttle variable
df$Throttle_Centered <- df$FullThrottlePct - mean(df$FullThrottlePct)
```

## modeling

### model 1 (baseline): effect of team status, engine, and year on relative performance gap (excluding very slow drivers)

```{r modeling}
# fit multiple regression model
model_v1 <- lm(PerToFastest ~ works + engine + Year, data = df_fast)

# get model summary
summary(model_v1)
cat("\n----------------\n")

# ANOVA table
car::Anova(model_v1, type = 2)
```

- model explains ~23% of the variation (Adj. R² = 0.2302)  
- model is statistically significant (F-test p < 0.001)
- main effects: works teams have a significantly smaller gap (~0.59 percentage points lower, p < 0.001), and recent years (2023–2025) show significantly smaller gaps compared to the other years
- engine differences: mercedes associated with a smaller gap (p < 0.01), renault associated with a larger gap (p < 0.01), and honda has a small negative effect that is statistically significant at the 5% level
- anova: works, engine, and year all have statistically significant effects

```{r check assumptions v1}
par(mfrow = c(2,2))
# get plots
plot(model_v1)     

# normality
shapiro.test(residuals(model_v1))    
cat("\n----------------\n")

# equal variance
car::leveneTest(PerToFastest ~ interaction(works, engine, Year), data = df_fast)
cat("\n----------------\n")

# independence
lmtest::dwtest(model_v1)

# VIF
vif(model_v1)
```

- residuals deviate from normality, evidence of heteroskedasticity (non-constant variance), evidence of positive autocorrelation (violates independence)

### model 2: sqrt transformation response to address variance issues

```{r}
# fit model 
model_v2 <- lm(sqrt(PerToFastest) ~ works + engine + Year, data = df_fast)

# get model summary
summary(model_v2)
cat("\n----------------\n")

# ANOVA table
car::Anova(model_v2, type = 2)
```

```{r check assumptions v2}
par(mfrow = c(2,2))
# get plots
plot(model_v2)

# normality
shapiro.test(residuals(model_v2))    
cat("\n----------------\n")

# equal variance
car::leveneTest(sqrt(PerToFastest) ~ interaction(works, engine, Year), data = df_fast)
cat("\n----------------\n")

# independence
lmtest::dwtest(model_v2)

# VIF
vif(model_v2)
```

- normality improved (residuals closer to normal), heteroskedasticity reduced but still present, positive autocorrelation remains (less severe)

### model 3: effect of team status, engine, and year on gap, controlling for throttle

```{r}
# fit model (add throttle as a covariate)
model_v3 <- lm(PerToFastest ~ works + engine + Year + FullThrottlePct, data = df_fast)

# model summary
summary(model_v3)
cat("\n----------------\n")

# ANCOVA table
car::Anova(model_v3, type = 2)

# Proportionality
test_interaction <- lm(PerToFastest ~ works * FullThrottlePct + engine * FullThrottlePct, data = df_fast)
anova(test_interaction)
```
- Adj. R² = 0.208 (not much change)
- FullThrottlePct is statistically significant (p < 0.001), with higher throttle usage associated with a slightly smaller performance gap

```{r check assumptions v3}
par(mfrow = c(2,2))
# get plots
plot(model_v3)     

# normality
shapiro.test(residuals(model_v3))    
cat("\n----------------\n")

# independence
lmtest::dwtest(model_v3)

# VIF
vif(model_v3)
```

### model 4 (interaction): year × work × engine effects on gap

```{r}
# fit model
model_v4 <- lm(PerToFastest ~ Year * works * engine, data = df_fast)

# model summary
summary(model_v4)
cat("\n----------------\n")

# ANOVA table
car::Anova(model_v4, type = 2)
```

- model fit improves (Adj. R² ≈ 0.40), and model is statistically significant overall (p < 0.001)
- main effects: works, engine, and year all have significant effects on performance
- interactions: all two-way interactions (year×works, year×engine, works×engine) and the three-way interaction (year×works×engine) is significant
- effect of team status and engine on performance gap varies over time, differences between engines and team types depend on year

```{r check assumptions v4}
par(mfrow = c(2,2))
# get plots
plot(model_v4)     

# normality
shapiro.test(residuals(model_v4))    
cat("\n----------------\n")

# independence
lmtest::dwtest(model_v4)
```


```

```{r pairwise contrasts - 3 way}
# pairwise engine comparisons within each year x work combination
emmeans(model_v4, pairwise ~ engine | Year * works, adjust = "bonferroni")
cat("\n----------------\n")

# pairwise works comparisons within each year x engine combination
emmeans(model_v4, pairwise ~ works | Year * engine, adjust = "bonferroni")
```

```{r pairwise contrasts - 2 way}
# create a temp model that does not use year
model_v4_noyear <- lm(PerToFastest ~ works * engine, data = df_fast)

# pairwise works comparisons within engine
emmeans(model_v4_noyear, pairwise ~ works | engine, adjust = "bonferroni")

# create a temp model that does not use engine
model_v4_noengine <- lm(PerToFastest ~ Year * works, data = df_fast)

# pairwise works comparisons within engine
emmeans(model_v4_noengine, pairwise ~ works | Year, adjust = "bonferroni")
```

### model comparison

```{r}
anova(model_v1, model_v3, model_v4)
```

- adding the covariate provides a statistically significant but small improvement but including interaction terms leads to a substantial improvement.